---
title: "Lab 4 team report"
author: "Isaac Tegeler"
date: "10/6/2020"
output: html_document
---


## Load Data
```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
defense <- read_excel('cyclonesFootball2019.xlsx', sheet='Defensive')
offense <- read_excel('cyclonesFootball2019.xlsx', sheet='Offensive')
biography <- read_excel('cyclonesFootball2019.xlsx', sheet='Biography')
```
# Part 1 Data Cleaning
## 1.1 Names as factors:
```{r}
defense$Name <- factor(defense$Name)
defense$Opponent_Opponent <-factor(defense$Opponent_Opponent)

offense$Name <- factor(offense$Name)
offense$Opponent_Opponent <-factor(offense$Opponent_Opponent)

biography$Name <- factor(biography$Name)
```
    
## 1.2 Convert stats to numeric type:
```{r}
offClean <- offense %>%
  separate(`Passing_CMP-ATT`, c("CMP", "ATT"), "-\\s+") %>%
  mutate(across(!Name & !Opponent_Opponent, as.numeric))

defClean <- defense %>%
  mutate_if(is.character , as.numeric)
```
    
## 1.3 Convert Height to inches
```{r}
bioClean <- biography %>%
  separate(Height, c("feet", "inches"), sep="-", convert=TRUE) %>%
  mutate(Height = (feet *12) + inches) %>%
  select(-feet, -inches)
```
## Part 1 results
```{r}
str(offClean)
str(defClean)
str(bioClean)
```


# Part 2 Tidying
## 2.1 Stat Column
```{r}
defStats <- defClean %>% pivot_longer(tail(colnames(defClean), 9), names_to="stat", values_to="count")
```

## 2.2 Rare Defesive Stats
```{r}
ggplot(defStats, aes(x = stat, y = count)) + geom_bar(stat="identity") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

The rarest skills for defense is turnovers.

## 2.3 Iowa State defense Vs. Iowa and Notre Dame

```{r}
x <- defStats %>%
  filter(stat=="Tackles_Solo") %>%
  filter(Opponent_Opponent == "Iowa" | Opponent_Opponent == "Notre Dame") %>%
  pivot_wider(names_from=Opponent_Opponent, values_from=count)


ggplot(x, aes(x=Iowa, y=`Notre Dame`)) + geom_point()
```

The plot shows that there are a lot more solo tackles for the players durring the Notre Dame game than the Iowa game. This idindcates the defense palyed better against Notre Dame.

## 2.4 City and State
```{r}
bioClean <- bioClean %>%
  separate(Hometown, c("City", "State"), sep=", ")

head(bioClean%>%select(City, State), 10)
```

## 2.5 Players by state
```{r}
ct <- bioClean %>%
  group_by(State) %>%
  summarise(players = n())

ct
```

# Part 3

## 3.1 players by state

```{r}
offenseStates <- offClean %>%
  select(Name) %>%
  inner_join(bioClean%>%select(Name, State), by="Name") %>%
  distinct()

deffenseStates <- defClean %>%
  select(Name) %>%
  inner_join(bioClean%>%select(Name, State), by="Name") %>%
  distinct()

offenseStates
deffenseStates
```

## 3.2 Brock Purdy vs ISU
```{r}

yearSum <- offClean %>%
  group_by(Name) %>%
  summarise_if(is.numeric, sum, na.rm = TRUE)

yearSum[is.na(yearSum)] <- 0
yatt <- yearSum %>%
  mutate(attempts = Rushing_ATT + Receiving_REC + ATT) %>%
  mutate(yards = Rushing_YDS *2 + Receiving_YDS + Passing_YDS) %>%
  mutate(ydsatt = yards/attempts)

result <- yatt %>% select(Name, ydsatt) %>% arrange(desc(ydsatt))
result
```
Based on the number of yards per touch the players on the team that the best are Milton and Shawn, while Purdy is nearly the worst.


## 3.3 2018 vs 2019

```{r}
defense18 <- read_excel('cyclonesFootball2018.xlsx', sheet='Defensive')
offense18 <- read_excel('cyclonesFootball2018.xlsx', sheet='Offensive')

defense18$Name <- factor(defense18$Name)
defense18$Opponent_Opponent <-factor(defense18$Opponent_Opponent)

offense18$Name <- factor(offense18$Name)
offense18$Opponent_Opponent <-factor(offense18$Opponent_Opponent)

offClean18 <- offense18 %>%
  separate(`Passing_CMP-ATT`, c("CMP", "ATT"), "-\\s+") %>%
  mutate(across(!Name & !Opponent_Opponent, as.numeric))

defClean18 <- defense18 %>%
  mutate_if(is.character , as.numeric)

yearSum18 <- offClean18 %>%
  group_by(Name) %>%
  summarise_if(is.numeric, sum, na.rm = TRUE)

yearSum18[is.na(yearSum18)] <- 0
yatt18 <- yearSum18 %>%
  mutate(attempts = Rushing_ATT + Receiving_REC + ATT) %>%
  mutate(yards = Rushing_YDS *2 + Receiving_YDS + Passing_YDS) %>%
  mutate(ydsatt = yards/attempts)


result_18 <- yatt18 %>% select(Name, ydsatt) %>% arrange(desc(ydsatt))
result_18

m <- full_join(result_18, result, by="Name", suffix=c("_18", "_19")) %>%
  mutate(diff = ydsatt_19-ydsatt_18) %>%
  na.omit() %>%
  select(Name, diff) %>%
  arrange(desc(diff))
m
```

The players that did not play both years were ommitted. The player that had the most improvement was MIlton and Nwangwu. While Purdy and Akers regressed from the previous season.
